---
output: github_document
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

# Exploring Trade Between Africa and other parts of the World

The goal of this repository is to use publicly available databases such as the World Bank's WITS Trade Stats to understand the flow of goods and services between African countries of interest and their largest trade partners. 


The Repo uses the **wbstats** package developed by World Bank developers to pull the latest data. 

## Load Required Libraries
```{r}
library(wbstats)
library(gt)
library(leaflet) # For plotting maps
library(tidyverse)
library(gganimate)
library(conflicted)
library(directlabels)

conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")

```

## Define Helper Functions
```{r, echo = TRUE}
#' Capitalize and make table titles bold
#'
#' @param indicator_id Indicator Id
#' @param indicator Indicator Name
#' @param source Database Source
#'
#' @return a formatted table with capitalized and bold titles.
#' @export
#'
#' @examples 
#' gt::gt() %>% prettify_table()
prettify_table <- function(indicator_id = NULL, indicator = NULL, source = NULL){
  
  labelled_table <- gt::gt() %>% 
    # Use clearer labels
     cols_label(indicator_id = "Indicator ID",
             indicator = "Indicator Name",
             source = "Source") %>% 
    # Make columns bold
    tab_style(
      locations = cells_column_labels(columns = everything()),
      style = list(cell_text(weight = "bold")
     )
   )

}


#' Title
#'
#' @param base_size 
#' @param base_family 
#' @param ticks 
#'
#' @return
#' @export
#'
#' @examples
theme_tufte_revised <- function(base_size = 11, base_family = "Gill Sans", ticks = TRUE) {
  
  ret <- ggplot2::theme_bw(base_family = base_family, base_size = base_size) +
    ggplot2::theme(
      axis.line = ggplot2::element_line(color = 'black'),
      axis.title.x = ggplot2::element_text(vjust = -0.3),
      axis.title.y = ggplot2::element_text(vjust = 0.8),
      legend.background = ggplot2::element_blank(),
      legend.key = ggplot2::element_blank(),
      legend.title = ggplot2::element_text(face="plain"),
      panel.background = ggplot2::element_blank(),
      panel.border = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank(),
      plot.background = ggplot2::element_blank(),
      strip.background = ggplot2::element_blank()
    )
  
  if (!ticks) {
    ret <- ret + ggplot2::theme(axis.ticks = ggplot2::element_blank())
  }
  
  ret
}

```

The World Bank tracks are over *200* Trade indicators with varying levels of missing data. In this repo, I only focus on indicators that are routinely used in macroeconomics publications. I will refer to both published economic literature and other trustworthy financial publications such as *Financial Times* and the *Wall Street Journal*.

```{r, include=TRUE, echo=FALSE}
trade_indicators <- wbstats::wb_indicators() %>%
  filter(grepl("Trade", indicator)) %>% 
  # The unit variable is empty(all NAs.). Exclude
  select(-unit) %>% 
  # Select aggregate metrics that are easity to understand. 
  # For example a metric as a % of GDP 
  filter(grepl("GDP", indicator, ignore.case = T))
```


```{r, include = FALSE, echo = TRUE, fig.cap = "Selected Indicators", fig.aligh = "center"}
trades_indicator_table <-
  trade_indicators %>% 
  select(indicator_id, indicator, source) %>% 
  gt::gt() #%>% 
  #prettify_table() 

# gt::gtsave(trades_indicator_table,
#            file = file.path(here::here(), "LNSHUTI.github.io/plots/trade_indicators.png"))
```

## A simple glance at one metrics

### Trade (% of GDP)

```{r, echo = TRUE, include = FALSE}
trade_pct_gdp <- 
  trade_indicators %>% 
  filter(indicator_id == "NE.TRD.GNFS.ZS") 
```

Pull data for trade as a % of GDP. The earliest available year is 1960. I suspect that most of the data before 2000 will be missing given that the Statistical Capacity of most countries was still very low and hence the data collection was not as reliable. 

The **wbstats** package includes a package that map each country's "iso3c" code to country name and continent.We pull this crosswalk and use it to label indicator data. 
```{r, echo = TRUE, include = TRUE}
df_trade <- wb_data(indicator = trade_pct_gdp$indicator_id) %>% 
  # Clean variable names
  janitor::clean_names() %>% 
  #Use Proper scale
  mutate(ne_trd_gnfs_zs = ne_trd_gnfs_zs*100) %>%
  select(iso3c, ne_trd_gnfs_zs, last_updated, date)

# Pull country labels
country_xwalk <- wb_countries() %>% 
  select(iso3c, region, income_level) %>% 
  unique()

# Add continent labels, joining by each country's isoc
df_tbl_labelled <- 
  df_trade %>% 
  left_join(country_xwalk, "iso3c") %>% 
  mutate(year = lubridate::year(paste0(date,"-01-01"))) %>%
  # Sum the metric by region-year
  group_by(region, year, year_labeller) %>% 
  summarize(median_trade_pct_gdp = median(ne_trd_gnfs_zs, na.rm = TRUE)/100) %>%
  ungroup()
 
# Plot the median of Trade as a % of GDP aggregated by region   
set.seed(125)
plt_trade_pct_gdp <- 
  df_tbl_labelled %>%
  ggplot(aes(x = year, y = median_trade_pct_gdp, color = region)) +
  geom_smooth(aes(x = year_labeller, y = median_trade_pct_gdp, color = region)) +
  geom_jitter(alpha = 0.01) +
  xlim(1958, 2022) +
  ylim(0, 100) +
  ylab("Trade as a % of GDP") +
  xlab("Year") +
  labs(color='Region') +
  #geom_vline(aes(xintercept = as.Date("2020-05-01")),lty=3) +
  scale_colour_brewer(palette = "Dark2") +
  ggthemes::theme_tufte() 

ggsave(plt_trade_pct_gdp, height = 4, width = 6,
       file = "LNSHUTI.github.io/plots/plt_trade_pct_gdp.png"
       )

plt_trade_pct_gdp_gif <-
  df_tbl_labelled %>%
  ggplot(aes(x = year, y = median_trade_pct_gdp, color = new_region)) +
  geom_point() +
  geom_jitter() +
  xlim(1958, 2022) +
  ylim(0, 100) +
  ylab("Trade as a % of GDP") +
  xlab("Year") +
  labs(color='Region') +

  theme_tufte_revised() +
  theme(legend.position="none", axis.text = element_text(size = 10)) +
  ylab("TRADE as a % of GDP") +
  #geom_vline(aes(xintercept = as.Date("2020-05-01")),lty=3) +
  scale_colour_brewer(name = "",palette = "Dark2", guide = FALSE)
  transition_time(year) 
  
animate(plot = plt_trade_pct_gdp_gif, height = 600, width = 400)
anim_save("LNSHUTI.github.io/plots/trade_by_region_year.gif")
```

# Trade Flows

The previous plot is interesting for two reasons. It shows what is widely known i.e that the global economy has been growing over the last few decades as shown by the overall growth in trade for all regions. The recession of 2008 and the contraction that followed the beginning of Covid-19 line up well with the plot. We can think of this step as a sanity check for ensuring that the data source is credible. 

Now to the interesting part. Let us use the same APIs to pull trade flow metrics from the World Bank database. We will link to an interactive shiny app that interested users can play with to examine trade flows between Africa and the rest of the world. <Link to Shiny APP>

Use the **wb_search()** function to search for all indicators that are related to trade. Display some of them and their corresponding descriptions. 

Of the trade metrics found, exclude those without a valid description. In addition, exclude metrics that are not broad enough to capture trends of interest. For example, indicators that are too sector specific such as **IC.FRM.TRD.TR1**, which stands for **Days to clear direct exports through customs** was excluded. It is beyond the scope of this notebook to go into firm-level comparisons.  

Finally, we limit ourselves to indicators that are measured in **US dollars.**  

```{r, max.height='100px'}
trade_metrics_tbl <- wb_search("TRD") %>% 
  # Exclude measures without a valid description
  filter(!is.na(indicator_desc)) %>% 
  # Exclude metrics that are too specific
  filter(!grepl(pattern = "FRM", x = indicator_id, ignore.case = T)) %>% 
  # Only include indicators denominated in US dollars
  filter(grepl(pattern = "dollar", x = indicator_id, ignore.case = T) | 
         grepl(pattern = "dollar", x = indicator_desc, ignore.case = T)) %>% 
  gt(data = .) %>% 
  cols_merge(
    columns = c(indicator_id, indicator),
    pattern = "{1}:\t{2}"
  ) %>% 
  cols_label(indicator_id = "Indicator ID", indicator_desc = "Detailed Description") %>% 
  # Given the large table size, only display some indicators and allow the user to scroll
  tab_options(container.overflow.y = T)
  
  
gtsave(data = trade_metrics_tbl,
      file = file.path(here::here(), "LNSHUTI.github.io/plots/global_trade_indicators.png"))
```

In addition to the World Bank data, we rely on **International Trade Data** from the Harvard data-verse. This source is richer and has more information related to the flow of goods and services between countries.  



**References:** 
Jesse Piburn (2020). wbstats: Programmatic Access to the World Bank API. Oak Ridge
  National Laboratory. Oak Ridge, Tennessee. URL
  https://doi.org/10.11578/dc.20171025.1827
  
John Graves(2019). health-care-markets: Defining Markets for Health Care Services.
  Vanderbilt University Medical Center. Nashville, TN. URL 
  https://github.com/graveja0/health-care-markets
  
The Growth Lab at Harvard University. International Trade Data (SITC, Rev. 2).
  2019-05-31. 2019. V5. Harvard Dataverse. URL. https://doi.org/10.7910/DVN/H8SFD2. 
  doi/10.7910/DVN/H8SFD2

Matt Grainger (2021). SDGsR: Interface with the UN SDGs API to get data about the 
  Sustainable Development Goals. R package version
  0.0.0.9000. https://drmattg.github.io/SDGsR/
  
  
https://tdhock.github.io/directlabels/HOCKING-useR-2011-directlabels-poster.pdf